Effectiveness of BNT162b2 booster doses in England: a observational study in OpenSAFELY

### 2022-02-28 manuscript/draft-manuscript.Rmd
Update draft-manuscript.Rmd

# 
# cmlinc <- read_csv(fs::path(output_dir_os, "cmlinc.csv")) %>%
#   mutate(
#     treatment_descr = fct_inorder(treatment_descr),
#     outcome_descr = fct_inorder(outcome_descr),
#     model_name = fct_inorder(str_wrap(model_name, 15)),
#     inc_CI_to = paste0(label_number(0.1, scale=1000)(1-survival.ul), " to ", label_number(0.1, scale=1000)(1-survival.ll))
#   )

@@

    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(label_number(accuracy=0.01)(hr), " (", label_number(accuracy=0.01)(hr.ll), "-", label_number(accuracy=0.01)(hr.ul), ")"),
    ve_display = paste0(label_percent(accuracy=0.1)(ve), " (", label_percent(accuracy=0.1, suffix="")(ve.ll), "-", label_percent(accuracy=0.1, suffix="")(ve.ul), ")"),
  )

@@

The UK COVID-19 vaccination programme delivered the first "booster" doses in mid September 2021, initially in groups at high risk of severe disease and then across the entire adult population. The BNT162b2 mRNA Pfizer-BioNTech and the Moderna mRNA-1273 vaccines were used.

@@

With the approval of NHS England we used the OpenSAFELY-TPP database, covering 40% of GP practices in England and linked to national coronavirus surveillance, hospital episodes, and death registry data, to estimate the effectiveness of a booster dose in "fully vaccinated" people. On each day of the recruitment period (`r format(as.Date(study_dates$studystart_date), "%d %B %Y")` to `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")`) each eligible person receiving a BNT162b2 booster dose on that day was recruited to the treatment group, and matched with an unboosted person recruited to the control group. Matching factors were region, JCVI priority group, and the day and type of the second vaccine dose. The treated and control groups across each day were pooled. Primary outcomes were documented SARS-CoV-2 infection, COVID-19-related accident and emergency attendance, unplanned COVID-19-related hospital admission, and COVID-19-related death occurring within 12 weeks of recruitment. Hazard ratios were estimated via a Cox model with additional covariates that were not used as matching factors.

@@

We matched `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` BNT162b2 booster recipients to unboosted controls with only two doses. During days 14-20 after the booster, the estimated hazard ratio (95% CI) comparing a BNT162b2 booster dose to two doses only was `r filter(hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for documented SARS-CoV-2 infection, `r filter(hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 death.

@@

BNT162b2 boosting was estimated to be highly effective in protecting against COVID-19 related hospitalisation and death.

@@

The national COVID-19 vaccination programme in England began a new phase on 16 September 2021 with the administration of the first booster COVID-19 vaccine doses in those who had already received their primary vaccination course (a completed first and second dose) (ref <https://www.england.nhs.uk/2021/09/nhs-begins-covid-19-booster-vaccination-campaign/>). Initially offered to those at highest risk of severe disease, eligibility was progressively extended and by 15 December 2021 every adult was eligible (<https://www.england.nhs.uk/2021/12/nhs-booster-bookings-open-to-every-eligible-adult/>). Booster doses were initially available no earlier than six months after dose two, but this was reduced to three months on 8 December 2021, following concerns over the surge in cases and the emergence of the Omicron variant (ref: <https://www.england.nhs.uk/2021/12/people-40-and-over-to-get-their-lifesaving-booster-jab-three-months-on-from-second-dose/>)

@@

We set out to estimate the effectiveness of BNT162b2 mRNA Pfizer-BioNTech booster doses versus two doses only in adults in England using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population. An analysis of effectiveness of Moderna mRNA-1273 boosters, which were administered from 29 October 2021, is planned when sufficient follow-up has accrued.

@@

The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at GP practices that use TPP SystmOne electronic health record software. It includes pseudonymized data such as coded diagnoses, medications and physiological parameters. No free text data are included. This primary care data is linked (via NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS). Healthcare worker status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

@@

We used observational data to emulate a sequence of target trials @dagan2021 @barda2021, starting on consecutive days between `r format(as.Date(study_dates$studystart_date), "%d %B %Y")` and `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")` inclusive.

Eligibility on each day was as follows: adults (18 years and over) who were registered at a GP practice using TPP's SystmOne clinical information system; received a primary COVID-19 vaccination course of either two BNT162b2 doses or two ChAdOx1-S doses (mRNA-1273 was not considered due to small numbers); not a vaccinated health care worker, not a resident in a care or nursing home, and not medically housebound or receiving end-of-life care (due to distinct clustering of vaccine uptake and infection risk in these groups that could not be reasonably measured and controlled for); no evidence of SARS-CoV-2 infection or COVID-19 disease within 90 days prior to recruitment; not undergoing an unplanned hospital admission; information on sex, ethnicity, deprivation, and NHS region was complete.

Additional eligibility criteria, based on sufficiently many people being vaccinated on each day by region, priority group, and week of second dose. If the rolling weekly average count of BNT162b2 booster doses within these strata was below 50 then recruitment did not occur on that day. This helped to ensure that unboosted people were both eligible and able to receive a booster dose at the time of potential recruitment.

We excluded matched pairs from days in which fewer than 10% of eligible treated people were successfully matched. See supplementary materials for further details.

## Treatment groups

For each trial, the treatment group comprised people who received a BNT162b2 booster dose on that day, and the control group comprised the same number of randomly selected unboosted people, matched on key characteristics as follows: primary course vaccine type (BNT162b2 or ChAdOx1-S); calendar week of the second vaccine dose; NHS region; prior infection; immunosuppression; vaccine priority group (as determined by the Joint Committee for Vaccine and Immunisation expert working group). People selected as controls were not eligible to be included again as a control in a subsequent trial, but were eligible for selection into the treatment group of a subsequent trial.

@@

The following three outcomes were considered: documented SARS-CoV-2 infection; COVID-19 hospital admission; and COVID-19 death. SARS-CoV-2 infections were identified using SGSS testing records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included, without differentiation between symptomatic and asymptomatic infection. COVID-19 hospital admissions were identified using HES in-patient hospital records with U07.1 or U07.2 reason for admission ICD-10 codes. COVID-19 deaths with the same COVID-19 ICD-10 codes mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death) were included.

@@

Each matched pair was followed from the day of recruitment (time zero or baseline) until the earliest of death, GP practice de-registration, booster receipt by the control, 10 weeks, or `r format(as.Date(study_dates$studyend_date), "%d %B %Y")`.

@@

For each outcome, we estimated cumulative incidence curves using the Kaplan-Meier estimator separately in the treated and control groups.

We estimated period-specific hazard ratios for treatment versus control in consecutive two-week periods using a Cox model. Baseline hazards were allowed to differ by recruitment day, region, and week and type of second vaccine dose. The following variables were included as covariates: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintile), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" classification, as per national prioritisation guidelines, the number of SARS-CoV-2 tests in the 6 months prior to the study start date (via SGSS), evidence of prior SARS-CoV-2 infection (positive SARS-CoV-2 test, "probable" infection documented in primary care, or COVID-19 hospitalisation), learning disabilities, severe mental illness, undergoing an unplanned hospital admission, undergoing a planned hospital admission. A robust variance estimator was used to account for participant clustering.

@@

We additionally estimated booster effectiveness in subgroups defined by primary vaccine course (both BNT162b2 or both ChAdOx1-S) by repeating the analysis separately in each subgroup.

@@

@@

After exclusions for missing values on demographic variables as described earlier, there were no missing values in remaining variables as they were each defined by the presence or absence of clinical codes or events in the patient record. This was therefore a complete-case analysis.

@@

A total of `r label_number(1, big.mark=",")(filter(matchflowchart, crit=="c0", treatment=="pfizer") %>% pull(n))` BNT162b2 booster recipients were identified between `r format(as.Date(matchsummary_treated$firstrecruitdate), "%d %B %Y")` and `r format(as.Date(matchsummary_treated$lastrecruitdate), "%d %B %Y")` in adults registered at a TPP practice, with `r label_number(1, big.mark=",")(filter(matchflowchart, treatment=="pfizer", crit=="c7") %>% pull(n))` (`r label_percent(0.1)(filter(matchflowchart, treatment=="pfizer", crit=="c7") %>% pull(pct_all))`) eligible for inclusion into the treatment group. Of these, `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_matched_eligible))`) were successfully matched with unboosted controls (Figure S1).

@@

Participant characteristics at trial recruitment between treatment groups (Table 1) were typically well-balanced though with some exceptions. For instance, unboosted controls had higher levels of deprivation, higher rates of learning disabilities and severe mental illness, and lower SARS-CoV-2 testing frequency.

There were `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(fup_years))` person-years of follow-up across both treatment groups (Table 2).

@@

Booster vaccination uptake over the duration of the study period, and the number of boosted people who were eligible and matched.

@@

xmin <- min(matchcoverage$vax3_date )
xmax <- max(matchcoverage$vax3_date )+1
matchcoverage  %>%
  filter(treatment=="pfizer") %>%
  mutate(
    # vax12_type = fct_case_when(
    #   vax12_type == "pfizer-pfizer" ~ "BNT162b2",
    #   vax12_type == "az-az" ~ "ChAdOx1-S",
    #   vax12_type == "moderna-moderna" ~ "mRNA-1273",
    #   TRUE ~ NA_character_
    # ),
    course = fct_case_when(
      vax12_type=="pfizer-pfizer" & treatment=="pfizer" ~ "2× BNT162b2  / 1× BNT162b2",
      vax12_type=="az-az" & treatment=="pfizer" ~ "2× ChAdOx1-S / 1× BNT162b2",
      vax12_type=="pfizer-pfizer" & treatment=="moderna" ~ "2× BNT162b2 / 1× mRNA-1273",
      vax12_type=="az-az" & treatment=="moderna" ~ "2× ChAdOx1-S / 1× mRNA-1273",
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot() +
  geom_col(
    aes(
      x=vax3_date+0.5,
      y=cumuln,
      group=fct_rev(status),
      fill=fct_rev(status),
      colour=NULL
    ),
    position=position_stack(),
    alpha=0.8,
    width=1
  )+
  geom_rect(xmin=xmin, xmax= xmax+1, ymin=0, ymax=6, fill="grey", colour="transparent")+
  facet_wrap(vars(course))+
  scale_x_date(
    breaks = unique(lubridate::ceiling_date(matchcoverage$vax3_date, "1 month")),
    limits = c(lubridate::floor_date((xmin), "1 month"), NA),
    labels = scales::label_date("%d/%m"),
    expand = expansion(add=1),
  )+
  scale_y_continuous(
    labels = scales::label_number(accuracy = 1, big.mark=","),
    expand = expansion(c(0, NA))
  )+
  scale_fill_brewer(type="qual", palette="Set2", direction = -1)+
  scale_colour_brewer(type="qual", palette="Set2", direction = -1)+
  labs(
    x="Date",
    y="Cumulative booster vaccines",
    colour=NULL,
    fill=NULL,
    alpha=NULL
  ) +
  theme_minimal()+
  theme(
    axis.line.x.bottom = element_line(),
    axis.text.x.top=element_text(hjust=0),
    strip.text.y.right = element_text(angle = 0),
    axis.ticks.x=element_line(),
    legend.position = "bottom"
  )+
  NULL
cat("  \n\n")
```

@@

Total follow-up and incidence rates for each outcome, by treatment group.

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath"),
    treatment == "pfizer"
  ) %>%
  left_join(
    matchsummary %>% select(treatment, n_eligible, prop_matched_eligible),
    by=c("treatment")
  ) %>%
  mutate(
    treatment_decr = paste0(
      treatment_descr, " - ",
      label_number(1, big.mark=",")(n_eligible), "eligible, ",
      label_number(1, big.mark=",")(n_1), "matched (",
      label_percent(0.1)(prop_matched_eligible), ")."
    )
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rrECI
  ) %>%
  mutate(
    rate_1=if_else(rate_1<0.001, "<0.001", label_number(0.001)(rate_1)),
    rate_0=if_else(rate_0<0.001, "<0.001", label_number(0.001)(rate_0)),
  ) %>%
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

Participants in the emulated trial contributed `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="postest", fup_period=="All") %>% pull(events))` documented infection, `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="covidadmitted", fup_period=="All") %>% pull(events))` COVID-19 hospital admissions, and `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="coviddeath", fup_period=="All") %>% pull(events))` COVID-19 deaths (Table 2).

@@

<!-- `r filter(hazards, subgroup=="none", outcome=="covidattendance", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 A&E attendances, `r filter(hazards, subgroup=="none", outcome=="covidcc", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 critical care admission,  -->

@@

For each outcome based on the fully adjusted model, the period-specific hazard ratios for boosting with BNT162b2 versus not boosting is shown, stratified by primary course. Models with less extensive confounder adjustment are provided in supplementary materials (supplementary Figure SX).

>>>>

km_none <- read_csv(fs::path(output_dir_os, "none", "km.csv"))  
km_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "km.csv"))

km <- 
  bind_rows(km_none, km_vax12_type) %>%
  mutate(
    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    outcome_descr_wrap = fct_relabel(outcome_descr, ~str_wrap(., width=20)),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
  )
# 
# cmlinc <- read_csv(fs::path(output_dir_os, "cmlinc.csv")) %>%
#   mutate(
#     treatment_descr = fct_inorder(treatment_descr),
#     outcome_descr = fct_inorder(outcome_descr),
#     model_name = fct_inorder(str_wrap(model_name, 15)),
#     inc_CI_to = paste0(label_number(0.1, scale=1000)(1-survival.ul), " to ", label_number(0.1, scale=1000)(1-survival.ll))
#   )

@@

    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    outcome_descr_wrap = fct_relabel(outcome_descr, ~str_wrap(., width=20)),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(label_number(accuracy=0.01)(hr), " (", label_number(accuracy=0.01)(hr.ll), "-", label_number(accuracy=0.01)(hr.ul), ")"),
    ve_display = paste0(label_percent(accuracy=0.1)(ve), " (", label_percent(accuracy=0.1, suffix="")(ve.ll), "-", label_percent(accuracy=0.1, suffix="")(ve.ul), ")"),
  )

metahazards_none <- read_csv(fs::path(output_dir_os, "none", "metaeffects.csv"))
metahazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "metaeffects.csv"))

metahazards <- 
  bind_rows(metahazards_none, metahazards_vax12_type) %>%
  mutate(

    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    outcome_descr_wrap = fct_relabel(outcome_descr, ~str_wrap(., width=20)),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(label_number(accuracy=0.01)(hr), " (", label_number(accuracy=0.01)(hr.ll), "-", label_number(accuracy=0.01)(hr.ul), ")"),
    ve_display = paste0(label_percent(accuracy=0.1)(ve), " (", label_percent(accuracy=0.1, suffix="")(ve.ll), "-", label_percent(accuracy=0.1, suffix="")(ve.ul), ")"),
  )

@@

The UK COVID-19 vaccination programme delivered the first "booster" doses in mid September 2021, initially in groups at high risk of severe disease and then across the adult population. The BNT162b2 mRNA Pfizer-BioNTech and the Moderna mRNA-1273 vaccines were used.

@@

With the approval of NHS England we used the OpenSAFELY-TPP database, covering 40% of GP practices in England and linked to national coronavirus surveillance, hospital episodes, and death registry data, to estimate the effectiveness of a booster dose versus two doses only. On each day of the recruitment period (`r format(as.Date(study_dates$studystart_date), "%d %B %Y")` to `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")`) each eligible person receiving a BNT162b2 booster dose was matched with an unboosted control who had received two vaccine doses. Primary outcomes were documented SARS-CoV-2 infection, COVID-19-related accident and emergency attendance, COVID-19-related hospital admission, and COVID-19-related death occurring within 10 weeks of recruitment. Hazard ratios were estimated using Cox models adjusting for additional covariates that were not used as matching factors. Follow-up spanned a period in which the Delta variant was predominant.

@@

We matched `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` BNT162b2 booster recipients to unboosted controls. During days 14-20 after the booster, the estimated hazard ratio (95% CI) comparing a BNT162b2 booster dose to two doses only was `r filter(hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for documented SARS-CoV-2 infection, `r filter(hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 death.

@@

We estimated high protection of BNT162b2 boosting against infection, COVID-19 hospitalisation, and COVID-19 death, though with some evidence of waning over time.

@@

The national COVID-19 vaccination programme in England began a new phase in September 2021 with administration of the first booster COVID-19 vaccine doses in those who had already received their primary vaccination course (a completed first and second dose) (ref <https://www.england.nhs.uk/2021/09/nhs-begins-covid-19-booster-vaccination-campaign/>). Initially offered to those at highest risk of severe disease, eligibility was progressively extended and by 15 December 2021 every adult in England was eligible (<https://www.england.nhs.uk/2021/12/nhs-booster-bookings-open-to-every-eligible-adult/>). Booster doses were initially available no earlier than six months after dose two, but this was reduced to three months on 8 December 2021, following concerns over the surge in cases and the emergence of the Omicron variant (ref: <https://www.england.nhs.uk/2021/12/people-40-and-over-to-get-their-lifesaving-booster-jab-three-months-on-from-second-dose/>). Vaccine prioritisation schedules were largely determined by recommendations from the Joint Committee for Vaccine and Immunisation (JCVI) expert working group.

In most cases, BNT162b2 and mRNA-1273 vaccines were used, and administered regardless of the vaccine brand used for the first and second doses. BNT162b2 was used from 16 September 2021, and mRNA-1273 from 29 October 2021 onwards.

@@

We estimated the effectiveness of booster vaccination with BNT162b2 versus two doses only in adults in England, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population. The study is conducted in an era where the Delta variant is dominant, with follow-up curtailed shortly after the emergence of Omicron.

@@

The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at English General Practices that use TPP SystmOne electronic health record software. It includes pseudonymized data such as coded diagnoses, medications and physiological parameters. No free text data are included. This primary care data is linked (via NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS).

@@

We used observational data to emulate a target trial @dagan2021 @barda2021 of the effectiveness of booster vaccination in people who had received two COVID-19 vaccine doses and were eligible for booster vaccination between `r format(as.Date(study_dates$studystart_date), "%d %B)` and `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")` inclusive.

Eligibility on each day was as follows: adults (18 years and over) who were registered at a GP practice using TPP's SystmOne clinical information system; received either two doses of BNT162b2 doses or two doses of ChAdOx1-S (mixed dosing and Moderna mRNA-1273 doses were not considered due to small numbers); not a health or social care worker, not a resident in a care or nursing home, and not medically housebound or receiving end-of-life care (due to distinct clustering of vaccine uptake and infection risk in these groups that could not be reasonably measured and controlled for); no evidence of SARS-CoV-2 infection or COVID-19 disease within the previous 90 days; not undergoing an unplanned hospital admission; information on sex, ethnicity, deprivation, and NHS region was complete.

Additionally, if the rolling weekly average count of BNT162b2 booster doses within strata defined by region, JCVI group, and week of second dose was below 50 then recruitment did not occur for that strata on that day. This helped to ensure that recruitment was restricted to those who were boosted in line with national prioritisation schedules and that the matched controls were both eligible and able to receive a booster dose at the time of potential recruitment.

The supplementary materials provide further details of the matching process.

## Matching and treatment groups

For each trial, the booster (treatment) group comprised people who received a BNT162b2 booster dose on that day, and the unboosted (control) group comprised the same number of randomly selected unboosted people, matched on the following characteristics: primary course vaccine type (BNT162b2 or ChAdOx1-S); date of the second vaccine dose, within 7 days; NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West); any evidence of prior SARS-CoV-2 infection (positive SARS-CoV-2 test, "probable" infection documented in primary care, or COVID-19 hospitalisation); Clinical risk group, defined by JCVI (clinically extremely vulnerable, clinically at-risk, neither); age, within 3 years, and by age groups defined by JCVI. Note that the matching and exclusion criteria jointly ensure that treatment groups are also matched within JCVI priority group.

People selected as controls were not eligible to be included again as a control in a subsequent trial, but were eligible for selection into the treatment group of a subsequent trial.

<!-- We excluded matched pairs from days in which fewer than 10% of eligible treated people were successfully. -->

@@

The following three outcomes were considered: documented SARS-CoV-2 infection; COVID-19 hospital admission; COVID-19 death; and non-COVID-19 death as a negative control outcome. SARS-CoV-2 infections were identified using SGSS testing records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included, without differentiation between symptomatic and asymptomatic infection. COVID-19 hospital admissions were identified using HES in-patient hospital records with U07.1 or U07.2 reason for admission ICD-10 codes. COVID-19 deaths with the same COVID-19 ICD-10 codes mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death) were included.

@@

Each matched pair was followed from the day of recruitment (i.e., time zero) until the earliest of death, GP practice de-registration, booster receipt by the control, 10 weeks, or `r format(as.Date(study_dates$studyend_date), "%d %B %Y")`.

@@

For each outcome, we estimated unadjusted cumulative incidence curves using the Kaplan-Meier estimator separately in the boosted and unboosted groups.

We use Cox models, stratified by recruitment day, NHS region, and brand of second vaccine dose, to estimate hazard ratios comparing boosted with unboosted people. We estimated effectveness over all follow-up time (0-70 days), and separately in days 0-6, 7-13, 14-27, 28-41, 42-70. Models were adjusted for additional variables not used as matching factors, as described below. A robust variance estimator was used to account for participant clustering.

@@

We additionally estimated booster effectiveness in the following subgroups: primary vaccine course (both BNT162b2 or both ChAdOx1-S); age (18-64 versus 65 and over); clinically extremely vulnerable. Analyses were conducted independently in each subgroup.

@@

## Covariates

The following variables were included as covariates: sex (male or female); English Index of Multiple Deprivation (IMD, grouped by quintile); ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census); morbidity count (diabetes, BMI > 40, chronic heart disease, chronic kidney disease, chronic liver disease, chronic respiratory disease or severe asthma, chronic neurological disease); learning disabilities; severe mental illness; the number of SARS-CoV-2 tests in the 6 months prior to the study start date; elective in-hospital episode at the time of recruitment.

The supplementary materials provide further details of the codelists and data sources used for all variables in the study.

@@

After exclusions for missing values on demographic variables, there were no missing values in remaining variables as they were each defined by the presence or absence of clinical codes or events in the patient record. This was therefore a complete-case analysis.

@@

`r label_number(1, big.mark=",")(filter(matchflowchart, crit=="c0", treatment=="pfizer") %>% pull(n))` BNT162b2 booster recipients were identified between `r format(as.Date(matchsummary_treated$firstrecruitdate), "%d %B %Y")` and `r format(as.Date(matchsummary_treated$lastrecruitdate), "%d %B %Y")` in adults registered at a TPP practice, with `r label_number(1, big.mark=",")(filter(matchflowchart, treatment=="pfizer", crit=="c7") %>% pull(n))` (`r label_percent(0.1)(filter(matchflowchart, treatment=="pfizer", crit=="c7") %>% pull(pct_all))`) eligible for inclusion into the treatment group. Of these, `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_matched_eligible))`) were successfully matched with unboosted controls (Table S1).

@@

Participant characteristics at trial recruitment between treatment groups (Table 1) were typically well-balanced though with some exceptions. For instance, unboosted controls had higher levels of deprivation, higher rates of learning disabilities and severe mental illness, and lower SARS-CoV-2 testing frequency, though the standardised mean difference was consistently below 0.1 (Supplementary Figure S1).

@@

Cumulative booster vaccination over the duration of the study period, by matching eligibility and matching success.

@@

xmin <- min(matchcoverage$vax3_date )
xmax <- max(matchcoverage$vax3_date )+1
data_coverage <- matchcoverage  %>%
  mutate(
    status = factor(status, levels=c("ineligible", "unmatched", "matched")),
    status_descr = fct_relabel(status, str_to_title)
  ) %>%
  arrange(status) %>%
  mutate(
    status_descr = fct_inorder(status_descr)
  ) %>%
  filter(treatment=="pfizer") %>%
  mutate(
    # vax12_type = fct_case_when(
    #   vax12_type == "pfizer-pfizer" ~ "BNT162b2",
    #   vax12_type == "az-az" ~ "ChAdOx1-S",
    #   vax12_type == "moderna-moderna" ~ "mRNA-1273",
    #   TRUE ~ NA_character_
    # ),
    course = fct_case_when(
      vax12_type=="pfizer-pfizer" & treatment=="pfizer" ~ "2× BNT162b2  / 1× BNT162b2",
      vax12_type=="az-az" & treatment=="pfizer" ~ "2× ChAdOx1-S / 1× BNT162b2",
      vax12_type=="pfizer-pfizer" & treatment=="moderna" ~ "2× BNT162b2 / 1× mRNA-1273",
      vax12_type=="az-az" & treatment=="moderna" ~ "2× ChAdOx1-S / 1× mRNA-1273",
      TRUE ~ NA_character_
    )
  )

data_annotate <- 
  data_coverage %>%
  filter(vax12_type=="pfizer-pfizer") %>%
  group_by(status_descr, course) %>%
  summarise(
    xpos = first(lubridate::floor_date((vax3_date), "1 month")),
    ypos= max(cumuln)
  ) %>%
  ungroup() %>%
  arrange(desc(status_descr)) %>%
  mutate(
    ypos=(cumsum(ypos) + lag(cumsum(ypos), 1, 0))/2
  ) %>%
  mutate(
    course="2× ChAdOx1-S / 1× BNT162b2"
  )

ggplot(data_coverage) +
  geom_col(
    aes(
      x=vax3_date+0.5,
      y=cumuln,
      group=status_descr,
      fill=status_descr,
      colour=NULL
    ),
    position=position_stack(),
    alpha=0.5,
    width=1
  )+
  geom_rect(xmin=xmin, xmax= xmax+1, ymin=0, ymax=6, fill="grey", colour="transparent")+
  geom_text(data=data_annotate, mapping=aes(x = xpos, y = (ypos), label = status_descr, colour=status_descr), hjust=0, show.legend = FALSE)+
  facet_wrap(vars(course))+
  scale_x_date(
    breaks = unique(lubridate::ceiling_date(matchcoverage$vax3_date, "1 month")),
    limits = c(lubridate::floor_date((xmin), "1 month"), NA),
    labels = scales::label_date("%d/%m"),
    expand = expansion(add=1),
  )+
  scale_y_continuous(
    labels = scales::label_number(accuracy = 1, big.mark=","),
    expand = expansion(c(0, NA))
  )+
  scale_fill_brewer(type="qual", palette="Set2", direction = -1)+
  scale_colour_brewer(type="qual", palette="Set2", direction = -1)+
  labs(
    x="Date",
    y="Cumulative booster vaccines",
    colour=NULL,
    fill=NULL,
    alpha=NULL
  ) +
  theme_minimal()+
  theme(
    axis.line.x.bottom = element_line(),
    axis.text.x.top=element_text(hjust=0),
    strip.text.y.right = element_text(angle = 0),
    axis.ticks.x=element_line(),
    legend.position = "none"
  )+
  NULL
cat("  \n\n")
```

@@

Total follow-up and incidence rates for each outcome, by treatment group. Cumulative incidences are provided in Supplementary Figure S2.

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath"),
    treatment == "pfizer"
  ) %>%
  left_join(
    matchsummary %>% select(treatment, n_eligible, prop_matched_eligible),
    by=c("treatment")
  ) %>%
  mutate(
    treatment_descr = paste0(
      treatment_descr, " - ",
      label_number(1, big.mark=",")(n_eligible), "eligible, ",
      label_number(1, big.mark=",")(n_1), "matched (",
      label_percent(0.1)(prop_matched_eligible), ")."
    )
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rrECI
  ) %>%
  mutate(
    rate_1=if_else(rate_1<0.001, "<0.001", label_number(0.001)(rate_1)),
    rate_0=if_else(rate_0<0.001, "<0.001", label_number(0.001)(rate_0)),
  ) %>%
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

Participants in the emulated trial experienced `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="postest", fup_period=="All") %>% pull(events))` documented infections, `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="covidadmitted", fup_period=="All") %>% pull(events))` COVID-19 hospital admissions, and `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="coviddeath", fup_period=="All") %>% pull(events))` COVID-19 deaths across `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(fup_years))` person-years of follow-up (Table 2).

@@

For days 42-70, estimated effectiveness (95% CI) comparing a BNT162b2 booster dose to two doses only, was `r filter(hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_left==42) %>% pull(ve_display)` for documented infection, `r filter(hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==42) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_left==42) %>% pull(ve_display)` for COVID-19 death (Figure 2), suggesting a potential waning of booster effectiveness with time. 
<!-- `r filter(hazards, subgroup=="none", outcome=="covidattendance", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 A&E attendances, `r filter(hazards, subgroup=="none", outcome=="covidcc", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 critical care admission,  -->

@@

For each outcome based on the fully adjusted model, the period-specific hazard ratios for boosting with BNT162b2 versus not boosting is shown, stratified by primary course. Models with less extensive confounder adjustment are provided in supplementary materials (Figure S3).
